Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?

### 2020-06-17 analysis/paper/paper.Rmd
Each sentence on new line, add references and introduction material

---
title: "Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?"
author: |
  | Adam H. Sparks^[Corresponding author: Adam.Sparks@usq.edu.au] \textsuperscript{1\dag}, Nancy P. Castilla \textsuperscript{1}, Bjoern Ole Sander \textsuperscript{1}
  | \textsuperscript{1}International Rice Research Institute,
  | Crop and Environmental Sciences Division, Los Baños, Laguna, Philippines, 4031
  | \textsuperscript{\dag} Current Address: University of Southern Queensland,
  | Centre for Crop Health, West Street,
  | Tooowoomba, Queensland, Australia, 4350
csl: "../templates/european-journal-of-plant-pathology.csl" # Insert path for the bib-style
abstract: |
  Water and nitrogen management play vital roles in rice production. However, the mismanagement of these two crop inputs may trigger the development of sheath blight of rice, caused by _Thanatephorus cucumeris_ (strain AG1-IA), which is favored by wet conditions, high relative humidity, and high nitrogen fertilizer levels. To understand how different combinations of water and nitrogen management affect sheath blight epidemics, we conducted two separate split--plot experiments with a water saving (alternate wetting and drying) regime and a traditional puddled regime combined with differing nitrogen treatments in the dry seasons of 2015 and 2016. Disease was scored using the same methodology in both experiments using a sheath blight assessment scale for field evaluation developed at the International Rice Research Institute to assess the severity on infected sheaths and leaves while sheath blight incidence on tillers were counted per hill. We were unable to detect any difference in the incidence of tiller sheath blight due to irrigation, tiller and leaf sheath blight did clearly differ statistically by irrigation treatment, but leaf sheath blight severity did not. Our findings indicate that farmers can adopt water saving technologies without risking increased sheath blight incidence. We suggest that further cross--disciplinary research in this area is warranted. All data generated from this project and code used for analysis are available as a reproducible research compendium from <https://github.com/openplantpathology/rice_awd_pests>.
bibliography: references.bib
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    #reference_docx: "../templates/template.docx"
header-includes:
  - \usepackage[T1]{fontenc}
  - \usepackage[utf8]{inputenc}
  - \DeclareUnicodeCharacter{2500}{\mbox{\vrule height2.2ptdepth-1.8ptwidth.5em}}
---

@@

@@

The AWD technology has been identified as promising climate smart practice for different rice growing regions that can stabilize rice production in water scarce areas as well as help reduce the carbon footprint of rice production. Various countries, *e.g.*, Bangladesh, Vietnam, Thailand and Philippines, plan to widely apply AWD to rice production [@MOEF2015]. However, a change like this in the water regime in rice fields on a large scale could trigger unintended effects, for example changes in disease intensity or severity.

While the effects of climate change directly on rice disease have been studied [@luo1998effects; @Duku2016] the effects of mitigation practices have not. And because of the well documented effects of N [@Slaton2003; @Tang2007] and the effects of alternating wet and dry periods [@Castilla1996] on sheath blight; we established field experiments to study what effects AWD irrigation technologies and N fertilisation in combination could have on sheath blight incidence and severity in tropical rice production.

@@

Irrigation in AWD plots was determined by the water level in plots, *i.e.*, when the water level reached 15&nbsp;cm below the soil surface irrigation water was applied to a level of 5&nbsp;cm. In puddled plots a standing water layer of 3--5&nbsp;cm was maintained throughout the growing season.

@@

## 2015 Experiment

@@

### The Interaction of Irrigation and N Management

@@

It should be noted that the levels of leaf severity remained low (< 1 \%) throughout the growing season in all treatments for both years. As AWD can only feasibly be implemented in the dry season. Therefor, the sheath blight disease levels in these studies may be lower than they would be if conducted in the wet season. However, sheath blight remains an issue in both seasons and the use of AWD would not appear to increase the disease but may be effective in reducing it in some cases.

@@

Climate change is predicted to decrease sheath blight risk in the Korean peninsula over the longer term. However, over the next 10 -- 20 years the risk is predicted to remain the same as from 2000 -- 2010 [@kim2015predicting]. Therefore, it is likely to remain an important diease in the near future with little resistance available in current varieties. However, based on these studies the use of AWD technologies would not appear to be a factor that increases the incidence or severity of rice sheath blight disease.

Readers may find a reproducible research compendium that recreates this paper, that includes all data generated from this project and code used for analysis as a reproducible research compendium from <https://github.com/openplantpathology/rice_awd_pests> [@Sparks2019].

@@

The authors wish to thank the IRRI Biometrics group for their input in the design of the experiments and Dr R. King and Mr D. Burrell of the University of Southern Queensland Statistical Consulting Unit for their insightful comments on the analysis of this paper. We would also like to thank D. Amozola, P. Escandor and P. Lapis for their technical support in the field. Lastly, we would like to remember our team mate Michael Noel, an integral part of our plant disease management team at IRRI and who helped design and managed the first season of the project but passed away in September 2016.

>>>>

---
title: "Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?"
author: |
  | Adam H. Sparks^[Corresponding author: Adam.Sparks@usq.edu.au] \textsuperscript{1\dag}, Nancy P. Castilla \textsuperscript{1}, Bjoern Ole Sander \textsuperscript{1}
  | \textsuperscript{1}International Rice Research Institute,
  | Crop and Environmental Sciences Division, Los Baños, Laguna, Philippines, 4031
  | \textsuperscript{\dag} Current Address: University of Southern Queensland,
  | Centre for Crop Health, West Street,
  | Tooowoomba, Queensland, Australia, 4350
csl: "../templates/european-journal-of-plant-pathology.csl" # Insert path for the bib-style
abstract: |
  Water and nitrogen management play vital roles in rice production. However, the mismanagement of these two crop inputs may trigger the development of sheath blight of rice, caused by _Thanatephorus cucumeris_ (strain AG1-IA), which is favored by wet conditions, high relative humidity, and high nitrogen fertilizer levels. To understand how different combinations of water and nitrogen management affect sheath blight epidemics, we conducted two separate split--plot experiments with a water saving (alternate wetting and drying) regime and a traditional puddled regime combined with differing nitrogen treatments in the dry seasons of 2015 and 2016. Disease was scored using the same methodology in both experiments using a sheath blight assessment scale for field evaluation developed at the International Rice Research Institute to assess the severity on infected sheaths and leaves while sheath blight incidence on tillers were counted per hill. We were unable to detect any difference in the incidence of tiller sheath blight due to irrigation, tiller and leaf sheath blight did clearly differ statistically by irrigation treatment, but leaf sheath blight severity did not. Our findings indicate that farmers can adopt water saving technologies without risking increased sheath blight incidence. We suggest that further cross--disciplinary research in this area is warranted. All data generated from this project and code used for analysis are available as a reproducible research compendium from <https://doi.org/>.
bibliography: references.bib
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    #reference_docx: "../templates/template.docx"
header-includes:
  - \usepackage[T1]{fontenc}
  - \usepackage[utf8]{inputenc}
  - \DeclareUnicodeCharacter{2500}{\mbox{\vrule height2.2ptdepth-1.8ptwidth.5em}}
---

@@

Climate change will bring changes to the environments in which our food crops are produced in [@Reyer2013].
Rice (*Oryza sativa*) is the most widely grown cereal crop in a diverse set of environments globally and will will have to adapt to changing environments where it is grown as a result.

Climate change will affect the diseases that occur in rice as well as the crop itself.
Many studies have been conducted *in silico* that examine how diseases may change with the climate.
Rice blast is one of the most frequently studied [@Luo1998; @Duku2016].
As well as

In-field studies have focused on the effects of temperature on bacterial blight [@Webb2010] and the effects of drought [@Dossa2017].
The effects of climate change on sheath blight have been examined by Kim et al. [-@Kim2015], where

@@

The AWD technology has been identified as promising climate smart practice for different rice growing regions that can stabilize rice production in water scarce areas as well as help reduce the carbon footprint of rice production.
Various countries, *e.g.* Bangladesh, Vietnam, Thailand and Philippines, plan to widely apply AWD to rice production [@MOEF2015].
However, a change like this in the water regime in rice fields on a large scale could trigger unintended effects, for example changes in disease intensity or severity.

While the effects of climate change directly on rice disease have been studied, the effects of mitigation practices have not.
And because of the well documented effects of N [@Slaton2003; @Tang2007] and the effects of alternating wet and dry periods [@Castilla1996] on sheath blight; we established field experiments to study what effects AWD irrigation technologies and N fertilisation in combination could have on sheath blight incidence and severity in tropical rice production.

@@

Irrigation in AWD plots was determined by the water level in plots, *i.e.* when the water level reached 15&nbsp;cm below the soil surface irrigation water was applied to a level of 5&nbsp;cm.
In puddled plots a standing water layer of 3--5&nbsp;cm was maintained throughout the growing season.

@@

@@

### The Interaction of Irrigation and N Management {-}

@@

It should be noted that the levels of leaf severity remained low (< 1 \%) throughout the growing season in all treatments for both years.
Therefore, the sheath blight disease levels in these studies may be lower than they would be if conducted in the wet season.
However, sheath blight remains an issue in both seasons and the use of AWD would not appear to increase the disease but may be effective in reducing it in some cases.

@@

Climate change is predicted to decrease sheath blight risk in the Korean peninsula over the longer term.
However, over the next 10 -- 20 years the risk is predicted to remain the same as from 2000 -- 2010 [@Kim2015].
Therefore, it is likely to remain an important diease in the near future with little resistance available in current varieties.
However, based on these studies the use of AWD technologies would not appear to be a factor that increases the incidence or severity of rice sheath blight disease.

Readers may find a reproducible research compendium that recreates this paper, that includes all data generated from this project and code used for analysis as a reproducible research compendium from <https://doi.org/...> [@Sparks2019].

@@

The authors wish to thank the IRRI Biometrics group for their input in the design of the experiments and Dr. R. King and Mr. D. Burrell of the University of Southern Queensland Statistical Consulting Unit for their insightful comments on the analysis of this paper.
We would also like to thank Mr. D. Amozola, Mr. P. Escandor and Mr. P. Lapis for their technical support in the field.
Lastly, we would like to remember our team mate Michael Noel, an integral part of our plant disease management team at IRRI and who helped design and managed the first season of the project but passed away in September 2016.
